Ages of Young Stars and Planets in the Kepler Field

### 2024-06-06 paper/ms.tex
update ms

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\newcommand{\berkeley}{Astronomy Department, University of California,
Berkeley, CA 94720, USA}
\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
Mellon University, Pittsburgh, PA 15213, USA}

@@

%\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
%\title{A Cliff In The Age Distribution of Young Stars and Planets in the Kepler Field}
\title{A Paucity of Young Stars and Planets in the Kepler Field}

\correspondingauthor{Luke G. Bouma}
\email{luke@astro.caltech.edu}

@@

@@

\author{and to be confirmed:}
\affiliation{\caltech}

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of sub-gigayear
  systems can
  be attributed to the star formation rate in the Kepler field
  dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend is observed both for known planet hosts as well
  as for the overall stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and it helps clarify the abundance
  of young and old planets that have been detected to date
  using transits.
  %
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets represent one means of building
the timeline for exoplanet evolution.  Informative individual
exemplars include V1298~Tau, a resonant chain of 
puffy planets that is a likely precusor to the compact
multiplanet systems \citep{David_2019}, and HIP~67522b, a Jupiter-sized
planet with a sub-Neptunian mass (\citealt{Rizzuto_2020}; Thao et
al.~submitted).  Population-level analyses have similarly suggested
differences in the size distribution of young exoplanets relative to
their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed from 0--10\,Gyr
\citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
clarifies the rarity of young stars: under this assumption,
$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
have $t$$<$1\,Gyr.  Studies of currently forming protoplanets
\citep{2018A&A...617A..44K}, and of exoplanets evolving just after
disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped
in their maximum achievable sample sizes by the tyranny of the
galactic star formation rate.

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
recent development, facilitated by Gaia, is the idea that finding
transiting planets is often easier than finding the host stellar
association \citep[e.g.][]{Tofflemire_2021}.

The ``cluster-first'' approach has a major limitation: most nearby
stars are in the field.  Only $\lesssim$1\% of stars
within $\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
The implication can appreciated by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Requiring say $t$$<$1\,Gyr at
2$\sigma$ precision gives $\approx$50 such planets at
the time of writing.  Most of these young planets are in clusters, and
were found by Kepler, K2, or TESS.  However, these surveys have
discovered $\approx$5{,}000 planets in total.  Assuming a constant
star formation history, we would expect an order of magnitude more
sub-gigayear transiting planets.

@@

Rotation-based ages have been reported for
various subsets of Kepler stars since the early data releases
\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
More recent work has further explored incorporating information from
stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
and from stellar colors, luminosities, and starspot amplitudes
\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
factors that can yield improvements, particularly for the ages of
sub-gigayear stars.   These factors are as follows.

\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018}.
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    This represents an improvement in the accuracy 
    of uncertainty propagation relative to previous calibrations.
  %
  \item We can identify binary stars with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer a sample of stars that we can use to test
    the reliability of the available single-star age-dating methods.
\end{enumerate}

Independently, recent years have also yielded improvements in the
lithium age scale.  Lithium ages include 
depletion boundary ages for M dwarfs and brown dwarfs in star
clusters, and decline-based ages for individual field
FGK stars \citep{Soderblom_2010}.  The latter
approach relies on the observed decline of Li abundances in
partially-convective stars as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who modeled the time evolution of the
\ion{Li}{1} 6708\,\AA\ equivalent width (EW) using a set of 6{,}200
stars in 52 open clusters.  Two-sided lithium ages are useful for
Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, though with a strong
dependence on spectral type.  The precision of lithium ages 
in this regime are now in the range of 0.3-0.5~dex.

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them in
Section~\ref{sec:litagecomparison} using
clusters in the Kepler field.  The population-level
trends are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of the stars with
  rotation periods are
  suitable for gyrochronology (dark gray points), based on factors
  including their non-binarity and proximity to the main
  sequence (orange line in lower panel; see Section~\ref{subsec:flags}).  Some
  host ``confirmed'' or ``candidate'' KOIs that meet additional
  planetary quality criteria (yellow points;
  Section~\ref{subsec:plflags}).  Surface gravities and effective
  temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  
    The gray lines are ``mean fits'' to the slow rotation sequences of
    open clusters.  The 2-D histogram of the stellar sample (left) includes only apparently
    single stars near the main sequence with $\log g$$>$4.2,
    RUWE$<$1.4, and temperatures of 3800--6200\,K.  
    The planet-hosts (right) require the same
    stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.  The number of stars $N_{\rm s}$ and
    planets $N_{\rm p}$are noted in 
    the relevant panels.
    A simple
    estimate for the number of young planets discovered by Kepler
    follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.  This
is not surprising, since the rotation periods of KOIs have received
considerably more scrutiny than those of ordinary Kepler stars.  We
therefore decided to split our subsequent analysis into a homogeneous
portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.  For the
latter portion, we first included 32 KOIs with orbital and rotation
periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the \citetalias{Santos_2019} and \citetalias{Santos_2021}
catalogs (A.~Santos, private communication).  We then incorporated an
additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013}, \citet{Mazeh_2015}
and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
periods is a supplementary measure aimed at completeness in our final
KOI age catalog; the provenances of the individual adopted periods are
noted in the relevant tables.

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK
stars in the open clusters used to
calibrate our gyrochronology model have $P_{\rm rot}$$\approx$35\,days.
Figure~\ref{fig:prot_vs_teff}
shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.

@@

We collected all blaze-corrected HIRES spectra from our group's
observations of non-false positive Kepler Objects of Interest with a
``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
least 10.  This yielded at least one spectrum for \nstwithspec\
stars hosting \nplwithspec\ planets.  About half of these stars have
measured rotation periods (\nstwithspecandprot\ stars and
\nplwithspecandprot, respectively).   For stars with multiple spectra
available, we analyzed only
the spectrum with the highest number of counts.  The resulting spectra
were acquired between 2009 September 6 and 2024 May 16, and
cumulatively comprise \nhireshours\,hours of open-shutter time.

@@

Our EW measurement approach did not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
precision of our method, after applying an initial
iteration on the \nlithiumstars\ KOIs with spectra, we compared our
lithium equivalent widths with those reported by
\citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
samples, we found broad agreement at $>$30\,m\AA, and significant
differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
required positive EWs, while we allowed for statistically negative ones.  At
$>$30\,m\AA, there is however a small offset in our respective scales caused
by our non-treatment of the iron blend,
such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.
We therefore directly subtracted
this constant mean value (7.50\,m\AA) when calculating lithium-based
ages.  This offset is at worst five times smaller than the astrophysical
scatter present in lithium EWs for calibration clusters at any given age
\citep[see][]{Jeffries_2023}.

@@

Our default source for stellar temperatures and surface gravities was
the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
\citep{2012AJ....144...24G}.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallaxes, and $V$-band extinction from the
\citet{2019ApJ...887...93G} reddening map.  The resulting stellar
parameters are available for \fracstarswithprotwithbtwenty\ of the
\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopted the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder are drawn
from DR25.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf The models.}
		Points represent $10^4$ draws from models that have been fitted to
		rotation periods \citep{Bouma_2023} and lithium equivalent widths
		\citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
		are the ``mean models'' at various ages.  The intrinsic dispersion
		in rotation and lithium about these mean models, which is what the
		models fit, sets the theoretical precision floor for the
		age-dating methods.    Additional sources of uncertainty,
		including measurement uncertainty, impose further limits on
		achievable precision.  These models are calibrated using open
		clusters younger than 4\,Gyr.  The displayed points assume a
		uniform distribution in temperature for visual clarity.  The sizes
		of the points are the same in each panel, so that low apparent
		density signifies greater dispersion around the mean.
		\label{fig:models}
	}
\end{figure*}

>>>>

%%%%%%%%%%%%%%%%
% INSTITUTIONS %
%%%%%%%%%%%%%%%%
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\newcommand{\berkeley}{Astronomy Department, University of California,
Berkeley, CA 94720, USA}
\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
Mellon University, Pittsburgh, PA 15213, USA}
\newcommand{\osaka}{Department of Earth and Space Science, Osaka University, Osaka 560-0043, Japan}

@@

%\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
%\title{A Cliff In The Age Distribution of Young Stars and Planets in the Kepler Field}
%\title{A Paucity of Young Stars and Planets in the Kepler Field}
\title{Ages of Young Stars and Planets in the Kepler Field}

\correspondingauthor{Luke G. Bouma}
\email{bouma.luke@gmail.com}

@@

%\author{and to be confirmed:}
%\affiliation{\caltech}

%  \author[0000-0003-1298-9699]{Kento Masuda} ?
%  \affiliation{\osaka}

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of sub-gigayear
  systems can
  be attributed to the star formation rate in the Kepler field
  dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  We observe this trend both for known planet hosts and also
  for the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets represent one means of building the
timeline for exoplanet evolution.  Informative individual exemplars
include HIP~67522b, a Jupiter-sized planet with a sub-Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
resonant chain of puffy planets that is a likely precusor to the
compact multiplanet systems \citep{David_2019}.  Population-level
analyses have similarly suggested differences in the size distribution
of young exoplanets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed from 0--10\,Gyr
\citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
suggests that
$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
have $t$$<$1\,Gyr.  Studies of currently forming protoplanets
\citep{2018A&A...617A..44K}, and of exoplanets evolving just after
disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped
in their maximum achievable sample sizes by the tyranny of the
galactic star formation rate.

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar, and
assumed planetary, ages are precise at the $\approx$10\% level.  A
recent development facilitated by Gaia is the idea that finding
transiting planets is often easier than finding the host star's birth
association \citep[e.g.][]{Tofflemire_2021}.

The ``cluster-first'' approach has a major limitation: most nearby
stars are in the field.  Only $\lesssim$1\% of stars
within $\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
The implication can appreciated by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Requiring $t$$<$1\,Gyr at
2$\sigma$ precision gives $\approx$50 such planets at
the time of writing.  Most of these young planets are in clusters, and
were found by Kepler, K2, or TESS.  However, these surveys have
discovered $\approx$5{,}000 planets in total.  Assuming a constant
star formation history, we would expect an order of magnitude more
sub-gigayear transiting planets.

@@

Rotation-based ages have been reported for
various subsets of Kepler stars since the early data releases
\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
More recent work has further explored incorporating information from
stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
and from stellar colors, luminosities, and starspot amplitudes
\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
factors that can yield improvements, particularly for 
sub-gigayear stars.   These factors are as follows.

\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018}.
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
  %
  \item Using open cluster data, we can marginalize over the range of
    ages that might explain any one star's rotation period
    \citep{Bouma_2023}.  This represents an improvement in the
    accuracy of uncertainty propagation relative to previous
    calibrations.
  %
  \item We can identify binary stars with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
    These clusters offer stars that we can use to test
    the reliability of the available single-star age-dating methods.
\end{enumerate}

Independently, recent years have also yielded improvements in the
lithium age scale.  Lithium ages include 
depletion boundary ages for M dwarfs and brown dwarfs in star
clusters, and decline-based ages for individual field
FGK stars \citep{Soderblom_2010}.  The latter
approach relies on the observed decline of Li abundances in
partially-convective stars as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
Empirical understanding however has improved due to the work by
\citet{Jeffries_2023}, who modeled the time evolution of the
\ion{Li}{1} 6708\,\AA\ equivalent width (EW) using a set of 6{,}200
stars in 52 open clusters.  Two-sided lithium ages are useful for
Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, though with a strong
dependence on spectral type.  The precision of lithium ages 
in this regime are now in the range of 0.3-0.5~dex.

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them in
Section~\ref{sec:litagecomparison} using
clusters in the Kepler field.  We discuss population-level
trends in Sections~\ref{sec:results} and~\ref{sec:disc}, and offer a
few conclusions in Section~\ref{sec:conclusions}.

@@

\begin{figure}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of the stars with
  rotation periods are
  suitable for gyrochronology (dark gray points), based on factors
  including their non-binarity and proximity to the main
  sequence (orange line in lower panel; see Section~\ref{subsec:flags}).  Some
  host ``confirmed'' or ``candidate'' KOIs that meet additional
  planetary quality criteria (yellow points;
  Section~\ref{subsec:plflags}).  Surface gravities and effective
  temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  
    The gray lines are ``mean fits'' to the rotation sequences of
    open clusters.  The 2-D histogram of the stellar sample (left) includes only apparently
    single stars near the main sequence with $\log g$$>$4.2,
    RUWE$<$1.4, and temperatures of 3800--6200\,K.  
    The planet-hosts (right) require the same
    stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.  The number of stars $N_{\rm s}$ and
    planets $N_{\rm p}$ are noted in 
    the relevant panels.
    A simple
    estimate for the number of young planets discovered by Kepler
    follows by counting points below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.  This
is not surprising, since the rotation periods of KOIs have received
more scrutiny than those of ordinary Kepler stars.  We
therefore decided to split our subsequent analysis into a homogeneous
portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.  For the
latter portion, we first included 32 KOIs with orbital and rotation
periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the \citetalias{Santos_2019} and \citetalias{Santos_2021}
catalogs (A.~Santos, private communication).  We then incorporated an
additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
\citet{David_2021} described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013}, \citet{Mazeh_2015}
and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
periods is a supplementary measure aimed at completeness in our final
KOI age catalog; the provenances of the individual adopted periods are
noted in the relevant tables.

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK
stars in the open clusters used to
calibrate our gyrochronology model have $P_{\rm rot}$$\approx$35\,days.
Figure~\ref{fig:prot_vs_teff}
shows a subset of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.

@@

We collected all blaze-corrected HIRES spectra from our group's
observations of non-false positive Kepler Objects of Interest with a
``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
least 10.  This yielded at least one spectrum for 1{,}464
stars hosting 2{,}174 planets.  About half of these stars have
measured rotation periods (\nstwithspecandprot\ stars and
\nplwithspecandprot\ planets, respectively).   For stars with multiple spectra
available, we analyzed only
the spectrum with the highest number of counts.  The resulting spectra
were acquired between 2009 September 6 and 2024 May 16, and
cumulatively comprise \nhireshours\,hours of open-shutter time.

@@

Our EW measurement approach did not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
precision of our method, after applying an initial
iteration on the \nlithiumstars\ stars with spectra, we compared our
lithium equivalent widths with those reported by
\citet{2018ApJ...855..115B}.  For the stars in both
samples, we found broad agreement at $>$30\,m\AA, and significant
differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
required positive EWs, while we allowed for statistically negative ones.  At
$>$30\,m\AA, there is however a small offset in our respective scales caused
by our non-treatment of the iron blend,
such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.
We therefore directly subtracted
this constant mean value (7.50\,m\AA) when calculating lithium-based
ages.  This offset is at worst five times smaller than the astrophysical
scatter present in lithium EWs for calibration clusters at any given age
\citep[see][]{Jeffries_2023}.

@@

Our default source for stellar temperatures and surface gravities was
the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
\citep{2012AJ....144...24G}.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
parallaxes, and $V$-band extinction from the
\citet{2019ApJ...887...93G} reddening map.  The resulting stellar
parameters are available for \fracstarswithprotwithbtwenty\ of the
\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopted the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which are
primarily derived from the photometric \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder are drawn
from DR25.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf The models.}
		Points represent $10^4$ draws from models that have been fitted to
		rotation periods \citep{Bouma_2023} and lithium equivalent widths
		\citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
		are the ``mean models'' at various ages.  The intrinsic dispersion
    around the mean, which is what the models fit, sets the
    theoretical precision floor for the
		age-dating methods.    Additional sources of uncertainty,
		including measurement uncertainty, impose further limits on
		achievable precision.  These models are calibrated using data from open
		clusters.  The sizes
		of the points are the same in each panel, so that low apparent
		density signifies greater dispersion around the mean.
		\label{fig:models}
	}
\end{figure*}
